---
title: "HW10 Forecast"
author: "Aaron"
date: "July 20, 2016"
output: 
  html_document: 
    highlight: haddock
    theme: cosmo
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r message=F, warning=F}

library(ggplot2)
library(ggfortify)
library(dplyr)
library(gridExtra)
library(forecast)

data("AirPassengers")
data("EuStockMarkets")

```

all the time series plot in this work will be **fortified** by `ggplot` and `ggfortify`.

`AirPassengers` and `EuStockMarkets` data are already `ts` object, so skip the `ts()` steps. 

### Using `decompose`

```{r}

AirPassengers %>% 
    decompose() %>% 
    autoplot()

```

```{r warning=F}

EuStockMarkets[,1] %>% 
    decompose() %>% 
    autoplot()

```

### Use `HoltWinters()` with different alpha and beta value 

```{r}

iter_hw = function(i, j, data=AirPassengers) {
    
   data %>% 
        HoltWinters(alpha = i, beta = j, gamma = NULL) %>% 
        autoplot() + 
        ggtitle(paste0("alpha = ", i, " | beta = ", j))
    
}

```

try different `alpha` with `beta` as null

```{r}

grid.arrange(iter_hw(0.25, NULL), iter_hw(0.5, NULL), iter_hw(0.75, NULL), iter_hw(1, NULL), 
             ncol = 2, nrow = 2)
```

We can try different beta too, same thing, this time with `EuStockMarkets` and alpha = 0.5

```{r}

grid.arrange(
    iter_hw(0.5, 0.25, EuStockMarkets[,1]), 
    iter_hw(0.5, 0.5, EuStockMarkets[,1]), 
    iter_hw(0.5, 0.75, EuStockMarkets[,1]), 
    iter_hw(0.5, 1, EuStockMarkets[,1]), 
    ncol = 2, nrow = 2)


```

### now try `predict()` with arima for better prediction

```{r}

AirPassengers %>% 
    auto.arima() %>% 
    forecast(level = c(95), h = 50) %>% 
    autoplot()

```

